---
author: "Will Tirone"
title: "Methods 6"
format: pdf
---

```{r include=FALSE}
library(tidyverse)
library(survey)
ncvs = read.csv('ncvs2000.csv')
```

# Q1) 

Here, we'll just multiply the successive probabilities of being selected together.

$$
\pi_{ij} = \frac{n x_i}{t_x} \frac{m_i x_{ij}}{x_i} \frac{q_{ij}}{x_{ij}} = \frac{n m q_{ij}}{t_x}
$$
# Q2)

First, we have to create unique values for the ppsu's, we'll do this by multiplying 
by 1000 and adding the ppsu's. 

```{r}
ncvs = ncvs |> 
  mutate(ppsu_unique = pstrat * 1000 + ppsu)

ncvs_survey = svydesign(~ppsu_unique, strata=~pstrat, weight=~pweight, data=ncvs)
```

## a) 

Here, I checked the codebook and `numinc` is the number of incidents per person, 
so we just use the usual code. The total is 15,850,360. 

```{r}
svytotal(~numinc, ncvs_survey)
```

## b)

Same thing here, just using `medexp` for medical expenses. The total is 
$86,285,440

```{r}
svytotal(~medexp, ncvs_survey)
```

## c) 

And a similar thing here but with robbery. The mean is fairly low, which probably 
makes sense. About 1 in 610 people commits a robbery. 

```{r}
svymean(~robbery, ncvs_survey)
```







