```{r}
library(tidyverse)
library(rvest)
library(polite)
library(stringr)
```

# Cost of the World Cup 

Reading: https://crossworks.holycross.edu/econ_working_papers/180/

```{r}

url = "https://en.wikipedia.org/wiki/Economics_of_the_FIFA_World_Cup"
bow = polite::bow(url)

# scrape the wiki table for costs 
out = 
  polite::scrape(bow) |> 
  html_nodes('table.wikitable') |> 
  html_table(fill = TRUE)

# regex steps to get numbers out 
# and convert to numeric 
hist_costs = out[[1]] |> 
  rename(host = "Host",
         cost = "General cost") |> 
  separate(cost, c(NA,"cost",NA), sep = '[$[]') |> 
  separate(cost, c("cost", NA), sep = '[bm]') |> 
  separate(cost, c("cost",NA), sep = '[(]') |> 
  separate(host, c("host", "year", NA), sep='[()]') |> # pull out year 
  mutate(across(where(is.character), str_trim)) |> # strips white space 
  transform(cost = as.numeric(cost)) |> 
  arrange(cost)

# USA has dollar value in millions, not billions, so manually adjusting 
# check another source though 
hist_costs[hist_costs["host"] == "USA", "cost"] = 
  hist_costs[hist_costs["host"] == "USA", "cost"] / 1000

# bar plot 
hist_costs |> 
  ggplot() + 
  geom_bar(aes(x=host, y=cost,fill=year),stat='identity') + 
  labs(title = "Cost of Hosting World Cup",
       y = "Cost in Billions (USD)",
       x = "Host Country") 
```

*"Despite the fact Qatar's World Cup will take place in November and December -- when the temperature is around 20 degrees Celsius; perfect for players and fans -- the SC has continued to develop cooling technologies in order to provide a strong tournament legacy.*

*Dr. Saud explained: "Our stadiums can be used 24/7, all year round, leaving a legacy for Qatar after the tournament -- and leaving no white elephants.""*

```{r}
# plot average temperature 

```

Cost Table of Stadiums, maybe compare to other stadiums

https://www.sportingnews.com/us/football/news/cost-world-cup-qatar-how-much-paid-fifa-most-expensive-2022/i69pi7uree5ctahcjyuzd9fn

Air Cooled Stadiums

https://www.fifa.com/tournaments/mens/worldcup/qatar2022/news/dr-cool-the-mastermind-behind-qatar-2022-s-air-cooled-stadiums

Resources:

https://www.r-bloggers.com/2021/07/politely-scraping-wikipedia-tables-2/
